Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2018-11-22 scagnostics.Rmd
updating index exploration section

The structured projection is in variables 5 and 6. Two samples sizes are used: $n=(100, 1000)$. All variables are standardized to have mean 0 and standard deviation 1.

@@

Figure \ref{fig:data} shows samples from each of the families, of the nuisance and structured pairs of variables. Table \ref{tab:indexTable} compares the PPIs for structured projections against those for nuisance variables, based on 100 simulated data sets of each type, using sample size 1000. The lower and upper show the 5th and 95th percentile of values. All indexes, except convex show distinctly higher values for the structured projections. The convex index considers the noise projections more interesting than the structured projections, which might be expected, but also provides an argument to use (1-convex) as the PPI.

@@

The family of data sets can be used to assess the PPIs on the desirable properties: smoothness, squintability, flexibility, rotation invariance, and speed. The procedures are as follows:

<!-- - Compare the index values for structured vs nuisance variables, by simulating many examples from the model families, and recording the 90% confidence intervals. This checks if the PPI can detect the structure, that it has distinct differences between the sets of values. 
-->
1. Use the tour to show the index values along an interpolation between pairs of nuisance variables, $x_1 - x_2$ view either to the $x_3 - x_4$. The result is ideally a smooth change in low values. This checks the smoothness property. 
2. Use the tour to show the index values along an interpolation between a pair of nuisance variables $x_1 - x_2$ and the structured pair of variables $x_5 - x_6$. This examines the squintability, and smoothness. If the function is smooth and slowly increases towards the structured projection, then the structure is visible from a distance. 
3. Ease of optimisation depends on having a smooth function, and this is tested using a guided tour, showing the optimisation path, starting from $x_1 - x_2$.  One index is optimised to show how effectively the maximum is attained, and the values for other PPIs is examined along the same path, which allows an assessment of similarity between PPIs. 
4.Rotation invariance is checked by computing PPIs on rotations of the structured projection. 
5. Varying the parameters for PPIs is used to examine the effect on value, smoothness, how flexible the index can be to find different structure and whether the index is fiddly to tune. Trace plots between noise and structured projections are used to examine this. 
6. Benchmarking on a range of sample sizes is used to assess the computational speed.

@@

## Index traces over tour sequence of interpolated nuisance projections
\label{sec:smooth}

@@

```{r plotV1V2toV3V4, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between two nuisance projections. All projections would be nuisance so the PPI are ideally low and smooth, with little difference between sample sizes (solid lines: $n=1000$; dashed: $n=100$). The scagnostic PPIs are noisy, as is TIC. Some indexes have distinct differences in values between sample sizes."}
fullResMelt <- fullRes %>%
  mutate(convex = 1-convex) %>% 
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMelt %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

@@

## Index traces over tour sequence between nuisance and structured projections
\label{sec:squintability}

@@

## Optimisation
\label{sec:guided}

@@

The guided tour combines optimisation with interpolating between pairs of planes. Target planes of the path are chosen by a random search method to find projection planes with high index values, which are then interpolated between. It operates roughly like simulated annealing that as one gets closer to an optimal value the random search range is reduced.

Given the behaviour of the various index measures under an interpolated tour path we next try to use the tourr implementation of the guided tour together with selected index functions to uncover the special views in datasets 2 and 3 with 1000 datapoints.

@@

>>>>

@@

The structured projection is in variables 5 and 6 ($x_5, x_6$). Two samples sizes are used: $n=(100, 1000)$. All variables are standardized to have mean 0 and standard deviation 1. Figure \ref{fig:data} shows samples from each of the families, of the nuisance and structured pairs of variables. Table \ref{tab:indexTable} compares the PPIs for structured projections against those for nuisance variables, based on 100 simulated data sets of each type, using sample size 1000. The lower and upper show the 5th and 95th percentile of values. All indexes, except convex show distinctly higher values for the structured projections. The convex index considers the noise projections more interesting than the structured projections, which might be expected, but also provides an argument to use (1-convex) as the PPI.

@@

The procedures for assessing PPI properties of smoothness, squintability, flexibility, rotation invariance, and speed are examined for samples from the family of data sets are as follows:

<!-- - Compare the index values for structured vs nuisance variables, by simulating many examples from the model families, and recording the 90% confidence intervals. This checks if the PPI can detect the structure, that it has distinct differences between the sets of values. 
-->
1. Compute the PPI values on the tour path along an interpolation between pairs of nuisance variables, $x_1 - x_2$ view either to the $x_3 - x_4$. The result is ideally a smooth change in low values. This checks the smoothness property. 
2. Chamge to a tour path between a pair of nuisance variables $x_1 - x_2$ and the structured pair of variables $x_5 - x_6$, and compute the PPI along this. This examines the squintability, and smoothness. If the function is smooth and slowly increases towards the structured projection, then the structure is visible from a distance. 
3. Use the guided tour to examine the ease of optimisation. This depends on having a relatovely smooth function, with structure visible from a distance. One index is optimised to show how effectively the maximum is attained, and the values for other PPIs is examined along the same path, to examine the similarity between PPIs. 
4. Rotation invariance is checked by computing PPIs on rotations of the structured projection. 
5. Some PPIs have a choice of parameters, and the choice can have an effect on function smoothness, and sensitivity to structure. The parameters for these PPIs is varied and examined along a tour path between noise and structured projections, to examine the effects of parameter choices and how fiddly the index is to tune. 
6. Computational speed for the selected indices is examined on a range of sample sizes.

@@

## PPI traces over a tour sequence of interpolated nuisance projections
\label{sec:smooth}

@@

```{r plotV1V2toV3V4, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between two nuisance projections. All projections would be nuisance so the PPI are ideally low and smooth, with little difference between sample sizes (solid lines: $n=1000$; dashed: $n=100$). The scagnostic PPIs are noisy. Some indexes have distinct differences in values between sample sizes."}
fullResMelt <- fullRes %>%
  mutate(convex = 1-convex) %>% 
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMelt %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

@@

## PPI traces over a tour sequence between nuisance and structured projections
\label{sec:squintability}

@@

## Optimisation check with the guided tour
\label{sec:guided}

@@

The guided tour combines optimisation with interpolation between pairs of planes. Target planes of the path are chosen to maximise the PPI. There are three derivative-free optimisation methods available in the guided tour: \verb# search_better_random# (1), \verb# search_better# (2), and \verb# search_geodesic# (3). Method 1 casts a wide net randomly generating projection planes, computing the PPIs and keeping the best projection. Method 2 conducts a localised maximum search. Method 3 is quite different: a local search is conducted to determine a promising direction, and then this direction is followed until the maximum in that direction is found. In each case, a single target plane is chosen by the search methods, and a tour path generated to interpolate from the current projection to the target plane.

@@

```{r eval=FALSE}
# Quick glimpse of optimisations
quartz()
animate_xy(pipeResc, guided_tour(mineIndex("TIC"), search_f = tourr:::search_better,
                                     cooling=1, alpha = 0.5, max.tries = 5000))
animate_xy(pipeResc, guided_tour(holes, search_f = tourr:::search_better,
                                     cooling=1, alpha = 0.5, max.tries = 5000))
```
